The Ky Fan Norms and Beyond: Dual Norms and Combinations for Matrix Optimization

### 2025-11-24 conf_article/neon.tex
Improved related work, added Ky Fan analysis

Motivated by the success of Muon, many generalizations and variations of it were proposed. Among the notable ones are Scion \citep{pethick2025training}, Dion \citep{ahn2025dioncommunicationefficientoptimizerlarge} and Gluon \citep{riabinin2025gluon}. Those researches target Muon's efficiency and establish convergence bounds. One central question, however, remains unanswered:

@@

\textbf{Schatten norms}
\citet{cesista2025schattenp} considered Schatten-$p$ norms $\norm{\mM^k}_{S_p} = \left(\sum_{i=1}^{\min(m, n)}\sigma_i^p\right)^{1/p}$. The update was as follows: $$X^{k+1} = X^k - \eta_k \mU \frac{\diag\left(\sigma_1^{q-1},\dots, \sigma_{\min(m,n)}^{q-1}\right)}{\norm{\mM^k}^{q-1}_{S_q}}\mV^\top$$ for $q$: $\frac{1}{p} + \frac{1}{q} = 1$. This formula recovers Neon when $p\rightarrow1$ provided that $\sigma_1 > \sigma_2$, which is true on real data; NSGD when $p=2$, and Muon when $p\rightarrow\infty$.

@@

\textbf{Ky Fan norms}
There is another family of matrix norms, which might help us: Ky Fan norms. For $k \in \{1, \dots, \min(m, n)\}$, the Ky Fan $k$-norm $\normkfk{\cdot}$ is $\sum_{i=1}^{k}\sigma_i$, i.e. the sum of the $k$ largest singular values of the matrix. Special cases of the Ky Fan $k$-norm are the Ky Fan $1$-norm, which is the sprectral norm, and the Ky Fan $\min\{m, n\}$-norm, which is the nuclear norm.

\textbf{Duals to Ky Fan norms}
While Schatten norms are closed under dualization and $\norm{\mM^k}_{S_p}^\dagger = \norm{\mM^k}_{S_q}$, for Ky Fan norms it is not generally the case: $\normkfk{\cdot}^\dagger = \max\{\frac{1}{k} \normn{\cdot}, \norms{\cdot}\}$ (see \citet{bhatia2013matrix}, p. 96). $k = 1$ and $k=\min(m, n)$ are exceptional: for $k=1$ the dual norm is $\normn{\cdot}$ with $k = \min(m, n)$ and for $k = \min(m, n)$ the dual norm is $\norms{\cdot}$ with $k=1$. For each other $k$ $\normkfk{\cdot} \neq \norm{\cdot}_{\text{KF-}k'}$ for any $k'$: $k'=\min(m, n)$ and $k'= 1$ correspond to the already discussed cases, while for other $k'$ one can change $\norm{\mM^k}_{\text{KF-}k'}$ by moving a small value between $\sigma_{k'}$ and $\sigma_{k'+1}$. During this change, $\norms{\mM^k}$ and $\normn{\mM^k}$ will remain constant, hence $\normkfk{\mM^k}$ will not change as well.

@@

	\begin{equation}\label{eq:ky_fan_update}
		\mX^{t+1} = \mX^{t} - \eta \sum_{i=1}^{k}u_i v_i^\top
	\end{equation}
\end{lemma}
In the section ~\cref{sec:matrix-side}, we will discuss how to compute the updates efficiently.

We introduce the family of {\it Fanions}, which consists of {\it Fanion-$k$}, lmo-based algorithms under the $\normkfk{\mM^t}^\dagger$ norms. By this terminology and the lemma, Muon is a Fanion-$\min\{n,m\}$, while Neon is an Fanion-1. Moreover, the centralized version of rank-$r$ Dion \citep{ahn2025dioncommunicationefficientoptimizerlarge} without the error feedback and without scaling of the update, from the perspective of lmo, is actually a Fanion-$r$.

Add more about Dion, and what its difference from us. I don't understand why their algorithm works well. I don't believe that the problem is with the error feedback.

\section{Algorithms for Matrices vs Algorithms for Vectors}
lmo optimizers in Schatten $S_p$ norms and in $l_p$ norms with common $p$ may be analogous to each other, as is illustrated by \cref{tbl:mat_vs_vec_lmo}. The analogies go beyond similarities in the updates. SignSGD is very close to Adam, as noted in \citet{bernstein2024oldoptimizernewnorm}, and both Adam and Muon perform well in training large models. NSGD is the same for both matrix and vector cases. Greedy Coordinate Descent methods are not applied to high-dimensional problems, from this perspective, it is not surprising that Neon underperforms on such problems.

@@

\section{Related Work}
As Muon \citep{jordan2024muon} is a very successful and popular optimizer for functions of weight matrices, a lot of reseach has been put into, first, further improving its performance, and, second, in explaining its success.

@@

	{\bf The nuclear norm in lmo.} As we found out only when writing this article, the nuclear norm has been already explored in the context of the linear minimization oracle. \citet{pethick2025sam} applied it to create $\nu$SAM, a new sharpness-aware minimization technique. The update from their Lemma 3.1 coincides with the update of Neon, but is used for completely different purposes. In addition, \citet{pethick2025sam} use power iterations to find $u_1$ and $v_1^\top$, while we suggest utilizing much more efficient and precise Lanczos algorithm.

	{\bf The Ky Fan Norm and Dion.} Rank-$k$ Centralized Dion, Algorithm 1 from \citet{ahn2025dioncommunicationefficientoptimizerlarge}, without an error feedback and scaling of the update, turns out to be an lmo-based algorithm under the $\normkfk{\cdot}^\dagger$ norm, which we described in {\it Different norms imply different updates}. Reported by the authors necessity of using error feedback to obtain satisfactory accuracy may take place in the cases of our algorithms as well, for example, F-Neon. This we leave to future research.

	{\bf Orthogonal research} Fanions, F-Fanions, and S-Fanions benefit from the general theoretical description of \citet{riabinin2025gluon}. They could be transformed to Drop-Fanions etc. (cite Drop-Muon article), they could be viewed as Broximal Point Methods (cite the article), they could be subject to EF (cite the EF for all). Of course, they could be clipped (cite ClipScion), and they could be analyzed when the LMO is not calculated accurately (cite Beyond the Ideal).

@@

@@

\section{Updates derivations}
% we prove a general lemma instead
%Proof of \cref{lemma:nsgd_muon_update}:
%Since $\norm{\cdot}^\dagger = \normfstar{\cdot}^{\dagger\dagger} = \normfstar{\cdot}$, the goal is to reach $\normfstar{\mM^k} = \alpha \tr{\mSigma} + (1-\alpha) \normf{\mM^k}$.

@@

Proof of \cref{lemma:ky_fan_update}:
Since $\norm{\cdot}^\dagger = \normkfk{\cdot}^{\dagger\dagger} = \normkfk{\cdot}$, the goal to reach $\normkfk{\mM^t}$.

Let us note that $\Delta = \sum_{i=1}^{k} u_i v_i^\top$ delivers the value. Indeed, $\<\mM^t, \Delta> =\<\mU \mSigma \mV^\top, \sum_{i=1}^{k} u_i v_i^\top> = \sum_{i,j=1}^{r, k}\<u_i \sigma_i v_i^\top, u_j v_j^\top> = \sum_{i=1}^{k}\sigma_i = \normkfk{\mM^t}$, which completes the proof.

>>>>

Motivated by the success of Muon, many generalizations and variations of it were proposed. Among the notable ones are Scion \citep{pethick2025training}, Dion \citep{ahn2025dion} and Gluon \citep{riabinin2025gluon}. Those researches target Muon's efficiency and establish convergence bounds. One central question, however, remains unanswered:

@@

\textbf{Schatten norms.}
\citet{cesista2025schattenp} considered Schatten-$p$ norms $\norm{\mM^k}_{S_p} = \left(\sum_{i=1}^{\min(m, n)}\sigma_i^p\right)^{1/p}$, which produce the update:
$$X^{k+1} = X^k - \eta_k \mU \frac{\diag\left(\sigma_1^{q-1},\dots, \sigma_{\min(m,n)}^{q-1}\right)}{\left(\sum_{i=1}^{\min(m,n)}{\sigma_i^q}\right)^{\frac{q-1}{q}}}\mV^\top$$
for $q$: $\frac{1}{p} + \frac{1}{q} = 1$. This formula recovers Neon when $p\rightarrow1$ provided that $\sigma_1 > \sigma_2$, which is true on real data; NSGD when $p=2$, and Muon when $p\rightarrow\infty$.

@@

\textbf{Ky Fan norms.}
There is another family of matrix norms, which might help us: Ky Fan norms. For $k \in \{1, \dots, \min(m, n)\}$, the Ky Fan $k$-norm $\normkfk{\cdot}$ is $\sum_{i=1}^{k}\sigma_i$, i.e. the sum of the $k$ largest singular values of the matrix. Special cases of the Ky Fan $k$-norm are the Ky Fan $1$-norm, which is the sprectral norm, and the Ky Fan $\min\{m, n\}$-norm, which is the nuclear norm.

Let us derive the update for an arbitrary $k$. Since $\normkfk{\cdot}^\dagger = \max\{\frac{1}{k} \normn{\cdot}, \norms{\cdot}\}$ (see \citet{bhatia2013matrix}, p. 96), the goal is to reach $\max\{\frac{1}{k}\sum_{i=1}^{\min(m,n)}\sigma_i, \sigma_1\}$. $\Delta = u_1 v_1^\top$ from Neon delivers $\sigma_1$, while $\Delta = \frac{1}{k}\sum_{i=1}^{\min(m,n)}u_i v_i^\top$ from Muon (with an extra factor of $1/k$) delivers $\frac{1}{k}\sum_{i=1}^{\min(m,n)}\sigma_i$. Thus, the update is either
$$\mX^{t+1} = \mX^{t} - \eta_t u_1 v_1^\top\text{ or }\mX^{t+1} = \mX^{t} - \frac{\eta_t}{k}\mU \mV^\top,$$
depending on which one better minimizes the function $F(\mX^t) + \mM^t(\mX^{t+1} - \mX^t)$.

The rank gap is not closed by the Ky Fan norms because the obtained update is either one-rank or full-rank.

\textbf{Duals to Ky Fan norms.}
While Schatten norms are closed under dualization and $\norm{\mM^k}_{S_p}^\dagger = \norm{\mM^k}_{S_q}$, for Ky Fan norms it is not generally the case. $k = 1$ and $k=\min(m, n)$ are exceptional: for $k=1$ the dual norm is $\normn{\cdot}$ with $k = \min(m, n)$ and for $k = \min(m, n)$ the dual norm is $\norms{\cdot}$ with $k=1$. For each other $k$, $\normkfk{\cdot} \neq \norm{\cdot}_{\text{KF-}k'}$ for any $k'$: $k'=\min(m, n)$ and $k'= 1$ correspond to the already discussed cases, while for other $k'$ one can change $\norm{\mM^k}_{\text{KF-}k'}$ by moving a small value between $\sigma_{k'}$ and $\sigma_{k'+1}$. During this change, $\norms{\mM^k}$ and $\normn{\mM^k}$ will remain constant, hence $\normkfk{\mM^k}$ will not change as well.

@@

	\begin{equation}\label{eq:ky_fan_update}
		\mX^{t+1} = \mX^{t} - \eta \sum_{i=1}^{k}u_i v_i^\top
	\end{equation}
\end{lemma}

\begin{proof}
Since $\norm{\cdot}^\dagger = \normkfk{\cdot}^{\dagger\dagger} = \normkfk{\cdot}$, the goal to reach $\normkfk{\mM^t}$.

Let us note that $\Delta = \sum_{i=1}^{k} u_i v_i^\top$ delivers the value. Indeed, $\<\mM^t, \Delta> =\<\mU \mSigma \mV^\top, \sum_{i=1}^{k} u_i v_i^\top> = \sum_{i,j=1}^{r, k}\<u_i \sigma_i v_i^\top, u_j v_j^\top> = \sum_{i=1}^{k}\sigma_i = \normkfk{\mM^t}$, which completes the proof.
\end{proof}
These updates constitute a family {\it Fanions}, LMO-based algorithms under the $\normkfk{\mM^t}^\dagger$ norms. The algorithm for a particular $k$ we will call {\it Fanion-$k$}. Thus, Muon is a Fanion-$\min\{m,n\}$, while Neon is a Fanion-1. Moreover, the unsharded version of the rank-$r$ Dion (Algorithm 1 from \citet{ahn2025dion}) without the error feedback and without scaling of the update is actually a Fanion-$r$ (see \citet{pethick2025understandingdion}, where Dion is written down in a notation more similar to ours).

@@

\section{Algorithms for Vectors $\rightarrow$ Algorithms for Matrices}

One cannot but notice the similarity between the updates LMO optimizers in the vector $l_p$ and the matrix Schatten $S_p$ norms, which is illustrated by \cref{tbl:mat_vs_vec_lmo}. However, the parallels are not limited to this similarity and can be traced even in the empirical observations. SignSGD is very close to Adam, as noted in \citet{bernstein2024oldoptimizernewnorm}, and both Adam and Muon perform well in training large models. NSGD is the same for both matrix and vector cases. Greedy Coordinate Descent is not applied to high-dimensional problems, so from this perspective it is not surprising that one-rank Neon underperforms on such problems.

\section{Related Work and Discussion}
As Muon \citep{jordan2024muon} is a very successful and popular optimizer for functions of weight matrices, a lot of reseach has been put into, first, further improving its performance, and, second, in explaining its success.

@@

	{\bf The LMO and Error Feedback.} We already mentioned that rank-$k$ unsharded Dion without error feedback and scaling of the update is Fanion-$k$. Since error feedback for Dion is very important as discovered by the ablation study in \citet{ahn2025dion}, F-Fanions and S-Fanions will benefit from it as well. In the context of federated learning, error feedback is effective even for compressed Muon \citep{gruntkowska2025efmuon}. The advantage of Fanions and S-Fanions is that less bits are required to transmit the updates: $\sum_{i=1}^k u_i v_i^\top$ is easily transmitted as $\{u_1, v_1, \dots, u_k, v_k\}$ ($(m+n)\times k$ floats), while the sign part is as usual coded in $m \times n$ bits. Thus, the compression is "built-in". Moreover, there is an intriguing possibility to construct differentially private Fanions and S-Fanions with the specific more optimal non-Gaussian noise, as was done with DP-signSGD \citep{jang2024dplogsign}. This we leave for future research.

	{\bf The nuclear norm in the LMO.} As we found out only when transforming our results into an article, the nuclear norm has already been explored in the context of the linear minimization oracle. \citet{pethick2025sam} applied it to create $\nu$SAM, a new sharpness-aware minimization technique. It would be interesting to substitute $\normn{\cdot}$ with $\normkfk{\cdot}^\dagger$ in their approach: $k > 1$ might increase accuracy boost, while preserving a small memory footprint and a small time overhead, if Dion-style power iterations are used.

	{\bf Orthogonal research} Fanions, F-Fanions, and S-Fanions benefit from the general theoretical description of \citet{riabinin2025gluon}. They could be transformed to Drop-Fanions by updating only some layers as in \citet{gruntkowska2025dropmuon}. They could be viewed as approximations of the Non-Euclidean Broximal Point Method for the corresponding norms \citep{gruntkowska2025noneuclideanbroximal}. One can clip them to produce ClippedScion-like algorithms \citep{pethick2025generalizedgradientnormclipping}. Finally, the results from \citet{shulgin2025inexactmuon} can be used to theoretically derive faster yet working approximate schemes to calculate the LMO; power iterations with a very limited number of iterations is a good thing to start with.

@@

We suggest that future works that deal with the non-Euclidean LMO explain in corollaries to their convergence theorems the empirical superiority of Muon to other Fanions. Without that, it is hard to believe that the bounds are relevant for practitioners.

@@

%\section{Updates derivations}
% we prove a general lemma instead
%Proof of \cref{lemma:nsgd_muon_update}:
%Since $\norm{\cdot}^\dagger = \normfstar{\cdot}^{\dagger\dagger} = \normfstar{\cdot}$, the goal is to reach $\normfstar{\mM^k} = \alpha \tr{\mSigma} + (1-\alpha) \normf{\mM^k}$.

@@
